---
title: "Predicting Flight Delays Between BWI and EWR: A Machine Learning Analysis of Weather, Network Effects, and Operational Factors"
date: 2026-03-13
categories: ["Machine Learning", "Aviation", "Flight Delays"]
repo-url: https://github.com/jonathanwilsonami/OR568_ML_Project
format:
  html:
    number-sections: true
    math:
      eq-prefix: "Model"
  pdf:
    number-sections: true
    keep-tex: true
authors:
  - "Jonathan Wilson"
  - "Karan [Last Name]"
  - "Irena [Last Name]"
  - "Val [Last Name]"
---

## Abstract

Flight delays remain one of the most persistent operational challenges in modern aviation, affecting airline efficiency, passenger satisfaction, and the broader transportation network. In this study, we examine flight delays within the BWI–EWR market pair using machine learning, exploratory data analysis, and engineered operational features.

Using Bureau of Transportation Statistics (BTS) flight records enriched with weather and temporal features, we construct predictors related to schedule timing, prior aircraft movement, weather conditions, and airport-level activity. We evaluate multiple supervised learning approaches including logistic regression, ridge regression, classification trees, random forests, and K-nearest neighbors.

Our findings suggest that delay behavior is shaped by a combination of operational and environmental effects, especially upstream aircraft delay propagation, airport congestion patterns, and weather-related variables. Tree-based approaches and penalized regression provide the most stable predictive performance, while simpler baseline models remain valuable for interpretation.

---

## Introduction & Background

### Motivation

Flight delays represent one of the most visible operational challenges in modern aviation. A single disruption can ripple across aircraft rotations, crews, passengers, and airports, creating cascading effects throughout the network.

Understanding the causes of delay propagation is critical for improving airline efficiency and passenger reliability.

### Related Works 

### Data and Methods
Aircraft registry features were initially considered, but because only a 2025 FAA registry snapshot was available while the flight dataset represents 2019 operations, these features could introduce temporal inconsistencies. Therefore, the analysis focuses primarily on operational variables derived directly from the flight dataset, such as aircraft rotation features, turnaround time, and previous-flight delays, which more directly capture delay propagation mechanisms.

- Add diagram of data pipeline and processing of features 
- Add diagram of model training and evaluation process

**Note: create a section of the site that represents reproducable instructions of our project and data processing steps, including code snippets and explanations.**

### Why the BWI–EWR Market Pair

The BWI–EWR corridor sits within one of the most operationally complex airspaces in the United States. Newark Liberty International Airport operates within the highly congested New York airspace, while BWI serves as a key mid-Atlantic hub.

Flights between these airports are vulnerable to multiple sources of disruption:

- upstream aircraft delays
- weather disruptions
- airspace congestion
- operational scheduling constraints

### What This Research Contributes

This study contributes a route-focused machine learning framework for airline delay analysis. By combining exploratory analysis, feature engineering, classification modeling, and model comparison, we provide a systematic approach for identifying the drivers of flight delays.

---

## Data

### Dataset Overview

This analysis uses flight operations data from the Bureau of Transportation Statistics (BTS). These records contain detailed flight-level information including:

- scheduled departure and arrival times
- actual departure and arrival times
- delay categories
- carrier information
- origin and destination airports

The dataset is enriched with weather variables and engineered temporal features.

### Limitations

Several limitations should be noted:

- BTS delay categories are aggregated and sometimes coarse
- weather observations may not perfectly represent local flight conditions
- route-level analysis cannot capture all national network effects
- airline operational decisions such as crew scheduling are not directly observed

Despite these limitations, the dataset provides a strong foundation for studying delay patterns.

---

## EDA  

- Show network map of airports and flight routes - some kind of data visualization that shows the network structure of the flights between BWI and EWR, as well as other connecting routes. This could be a graph or a map visualization that highlights the frequency of flights and the connections between different airports.

## Feature Engineering

### Delay Indicator

To simplify classification, we define a binary delay variable:

$$
Delayed_i =
\begin{cases}
1 & \text{if DepartureDelay}_i > 15 \\
0 & \text{otherwise}
\end{cases}
$$

### Prior Aircraft Delay

Upstream delays are captured by the delay of the previous aircraft flight:

$$
PriorDelay_i = ArrivalDelay_{previous}
$$

### Turnaround Time

Turnaround time captures the buffer between flights:

$$
Turnaround_i = DepartureTime_i - PreviousArrivalTime_i
$$

### Weather Severity

Weather features combine several meteorological indicators:

$$
WeatherSeverity = f(wind, visibility, precipitation, storm\ indicators)
$$

### Temporal Features

Time-based indicators include:

- departure hour
- day of week
- month
- holiday flags

These help capture cyclical delay patterns.

---

## Exploratory Data Analysis

### Distribution of Delays

```{r}
# Histogram of departure delays
# ggplot(df, aes(x = DepDelayMinutes)) +
#   geom_histogram(bins = 50)
```

### Delays by Time of Day

```{r}
# Delay frequency by departure hour
```

### Weather vs Delay

```{r}
# Weather severity vs delay outcome
```

### Delay Propagation

```{r}
# Previous arrival delay vs next departure delay
```

---

## Research Questions

This study seeks to answer the following questions:

1. Which operational and weather variables most strongly predict flight delays?
2. How strongly do delays propagate from prior flights?
3. Which machine learning models best classify delayed flights?
4. Can unsupervised methods reveal underlying structure in delay patterns?

---

## Building Classifiers

### Logistic Regression

We begin with a baseline logistic regression model.

$$
P(Y_i = 1|X_i) =
\frac{1}{1 + e^{-\eta_i}}
$$

$$
\eta_i = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + ... + \beta_p X_p
$$

```{r}
# logistic regression model
```

---

### Ridge Logistic Regression

To address multicollinearity and improve stability we apply ridge regularization.

```{r}
# ridge logistic regression
```

---

### Classification Tree

Decision trees provide interpretable classification rules.

```{r}
# classification tree
```

---

### Random Forest

Random forests improve predictive accuracy through ensemble learning.

```{r}
# random forest model
```

---

### K-Nearest Neighbors

KNN classifies flights based on nearby observations in feature space.

```{r}
# knn model
```

---

## Unsupervised Analysis

### Principal Component Analysis

PCA identifies dominant dimensions in the feature space.

```{r}
# PCA analysis
```

### K-Means Clustering

Clustering reveals natural groupings among flight observations.

```{r}
# kmeans clustering
```

---

## Model Comparison

We compare models using metrics such as:

- accuracy
- precision
- recall
- F1 score

```{r}
# model comparison table
```

---

## Results

### Key Findings

The analysis suggests that delay outcomes are influenced by several interacting factors:

- upstream aircraft delays
- airport congestion patterns
- weather conditions
- temporal scheduling effects

### Operational Interpretation

Predictive models can help identify flights at elevated risk of delay before departure, enabling airlines and planners to anticipate disruptions.

---

## Discussion & Limitations

Several limitations should be considered:

- missing operational factors such as crew and maintenance constraints
- imperfect weather measurements
- route-specific findings that may not generalize

Despite these challenges, the framework provides a replicable approach for studying delay behavior.

---

## Conclusion

This study demonstrates how machine learning methods can be applied to understand delay dynamics in the BWI–EWR market pair.

Tree-based methods and penalized regression provide the best balance of predictive performance and interpretability. The analysis highlights the importance of upstream delay propagation and temporal operational patterns.

Future work could extend this framework to larger airline networks and incorporate time-series modeling approaches.

---

## References


## Appendix  

### Data Catalog 


## Hardware 

The models were trained on a high-performance local workstation running Ubuntu 22.04.5 LTS (64-bit) with kernel version 6.8. The system is equipped with an AMD Ryzen 9 7900X processor featuring 24 logical cores and boost speeds up to 5.65 GHz, along with approximately 96 GB of system memory. The machine also includes both integrated AMD graphics and a discrete NVIDIA GPU, although model training was conducted using CPU resources. This hardware configuration provides substantial parallel processing capability and memory bandwidth, enabling efficient handling of large-scale data processing, feature engineering, and machine learning model training workflows.

## Models 

We are interested in building an multi-output model that both classifies an arrival delay and determins how long that arrival delay will be using 
features that are available during the present time. 

Models: 
LSTM -

Model Arch 
Break data into a step matrix - each example is broken into ordered steps in time or sequence, rather than treated as one flat row. The 
LSTM model needs this 3D matrix to work on. 
We apply some scaling to the data to standardize each feature this allows our neural network to train more effectively thus improving stability 
and convergence. 

Flights contexts in 2 hop would be 3 different time steps. t-2, t-1, t0. One sample would be 3 rows. 

Varients - Use same training function but different inputs and model sizes. 

Making the model bigger adds more context: Answers Does adding richer operational context improve sequence-based delay prediction? 

By keeping the layers to only 2 hops we prevent the exploding gradient problem 
To prevent the vanishing gradient problem we....else we would never reach our optimal value 
LSTMs help us avoid the vanishing and exploding gradient problems by using two seperate paths 
that are for long and short term memories. The gates helps us make relavant predictions 
only remmebering what is needed and forgetting what is not. (explain) 

Regularlization - We applied a dropout regularization to the LSTM outputs, randomly turning off a fraction (dropout, e.g., 20%) of the neurons during training. This helps prevent the model from overfitting by forcing it to not rely too heavily on any single pattern or feature.

Loss Functions 
Binary Cross-Entroy 

$$
\mathcal{L}_{BCE} = -\frac{1}{N} \sum_{i=1}^{N} \left[ y_i \log(p_i) + (1 - y_i) \log(1 - p_i) \right]
$$

MSE 
$$
\mathcal{L}_{MSE} = \frac{1}{N} \sum_{i=1}^{N} (y_i - \hat{y}_i)^2
$$

Evaluation Metrics 

Cross Validation and Hyper Parameter Tuning 

Train-Test Splits for CV. Expanding window folds inside Jan-Oct training period.

    Fold 1: train Jan-Jun, validate Jul
    Fold 2: train Jan-Jul, validate Aug
    Fold 3: train Jan-Aug, validate Sep
    Fold 4: train Jan-Sep, validate Oct
    
FINAL TEST SPLIT
#    Keep Nov-Dec untouched


Tree - 

Metrics:
Classification - AUC, F1, precision, recall, accuracy  
Regression - MAE, RMSE 

- Explain why used and are there better ones to use 

Rows in final modeling table: 6594553
Train rows: 5490189 (Jan 1-Oct 31)
Test rows : 1104364 (Nov 1-Dec 31)

The code performs a time-based train/test split on the flights dataset using November 1, 2019 as the cutoff. A temporal split like this is the correct approach for LSTM modeling on time-series data, as it mirrors real-world prediction conditions and prevents data leakage from future records into training.it is important to split your training and test sets by a fixed date rather than randomly. Random splitting allows future data to leak into the training set, meaning the model inadvertently learns from information it would not have had access to in a real-world scenario. By splitting on a specific date like November 1, 2019, we ensure the model is only trained on the past and evaluated on truly unseen future data, which gives a more honest and reliable measure of model performance.




