---
title: "Predicting Flight Delays Between BWI and EWR"
subtitle: "Machine Learning Analysis of Weather and Operational Effects"
author: 
  - Jonathan Wilson
  - Karan [Last Name]
  - Irena [Last Name]
  - Val [Last Name]
date: today

format:
  revealjs:
    theme: dark
    slide-number: true
    transition: slide
    code-fold: true

execute:
  echo: false
---

## Motivation

- Flight delays ripple through airline networks
- Small disruptions propagate across aircraft rotations
- Understanding delay drivers helps improve operational planning

Focus of this study:

**BWI → EWR market pair**

---

## Research Questions

1. What factors most strongly predict flight delays?
2. How much do delays propagate from earlier flights?
3. Which machine learning models perform best?
4. What structural patterns exist in the data?

---

## Data Sources

Primary dataset:

**Bureau of Transportation Statistics (BTS)**

Key variables:

- departure delays
- arrival delays
- carrier information
- route identifiers
- weather conditions
- time-of-day indicators

---

## Feature Engineering

Key engineered features:

**Delay Indicator**

$$
Delayed =
\begin{cases}
1 & Delay > 15 \\
0 & otherwise
\end{cases}
$$

**Prior Aircraft Delay**

$$
PriorDelay = ArrivalDelay_{previous}
$$

**Turnaround Time**

$$
Turnaround = DepartureTime - PreviousArrival
$$

---

## Exploratory Analysis

Important patterns we examine:

- distribution of delays
- time-of-day effects
- weather vs delay relationships
- delay propagation patterns

```{r}
# placeholder for EDA plots
```

---

## Modeling Approaches

We evaluate multiple models:

- Logistic Regression
- Ridge Logistic Regression
- Decision Trees
- Random Forest
- K-Nearest Neighbors

Each model is evaluated using:

- accuracy
- confusion matrix
- interpretability

---

## Logistic Regression

Baseline probabilistic model:

$$
P(Y=1|X) = \frac{1}{1+e^{-\eta}}
$$

$$
\eta = \beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_pX_p
$$

---

## Tree-Based Models

Decision trees provide interpretable rules.

Random forests improve performance using:

- ensemble learning
- bootstrap sampling
- feature randomness

```{r}
# tree visualization placeholder
```

---

## Model Comparison

Metrics used:

- Accuracy
- Precision
- Recall
- F1 Score

```{r}
# model comparison table
```

---

## Key Findings

Important drivers of delay:

- upstream aircraft delays
- time-of-day scheduling effects
- weather conditions
- airport congestion

Tree-based models produced the most reliable predictions.

---

## Operational Insights

Machine learning models can help:

- identify high-risk flights
- anticipate delay propagation
- improve scheduling decisions
- support airline operations analysis

---

## Limitations

- incomplete operational data
- weather approximation
- route-specific analysis
- limited network context

---

## Future Work

Potential improvements:

- expand to full airline network
- incorporate time-series models
- integrate air traffic control data
- include crew and maintenance data

---

## Conclusion

This study demonstrates that:

- delay propagation is a key driver of disruption
- machine learning models can effectively classify delays
- route-level analysis provides actionable operational insight

---

## Questions