---
title: "BWI–NY Market Pair"
date: 2026-03-13
format: html 
toc: true 
code-fold: false 
code-tools: true
---

# Overview
This modeling and analysis focuses on a specific market pair between BWI and the New York area to provide a more detailed, localized view of flight delay behavior. By narrowing the scope, we can more closely examine the factors that influence delays within a constrained and highly trafficked corridor.

This setting allows for deeper exploration of temporal patterns, weather effects, and aircraft rotation dynamics without the added complexity of the full national network. It also provides a controlled environment to test feature engineering approaches and modeling techniques.

The insights gained from this focused analysis serve as a foundation for understanding delay behavior, which can then be extended and generalized to the broader U.S. domestic network. See the page on U.S. Domestic Network Modeling for more details. 

## Setups 

The setup below loads all required packages via `load_project_packages()`, defined in `utils.r`, 
which installs any missing packages and attaches them to the session. Flight data is loaded from 
the project's data source and cached locally as a parquet file at `data/flights_raw.parquet` on 
first run — subsequent renders skip the 30-40 second load and read directly from the local cache 
instead. To refresh the data, delete `data/flights_raw.parquet` and re-render. `flights_raw` is 
returned as an Arrow lazy frame, meaning no data is held in memory until explicitly collected 
downstream.

```{r}
#| label: install-packages
#| message: false
#| warning: false

source("shared-notebooks/common_utils/r/utils.r") # common utils
load_project_packages()
set.seed(42)
```

```{r}
#| label: load-raw-data
#| message: false
#| warning: false

if (!file.exists("data/flights_raw.parquet")) {
  message("No cache found, loading and writing parquet...")
  load_flight_data() %>% 
    collect() %>% 
    janitor::clean_names() %>%
    arrow::write_parquet("data/flights_raw.parquet")
  message("Cache written to data/flights_raw.parquet")
}

flights_raw <- arrow::open_dataset("data/flights_raw.parquet")

```

# BWI-NY Modeling & Analysis 

## Holiday Analysis

```{r}
#| label: build-df-final
#| message: false
#| dependson: "load-raw-data"

# Define 2019 Federal Holidays
us_holidays_2019 <- as.Date(c(
  "2019-01-01", "2019-01-21", "2019-02-18", "2019-05-27",
  "2019-07-04", "2019-09-02", "2019-10-14", "2019-11-11",
  "2019-11-28", "2019-12-25"
))

# Generate travel windows (+/- 3 days around each holiday)
holiday_windows <- map(us_holidays_2019, ~ seq(.x - 3, .x + 3, by = "day")) %>%
  flatten_dbl() %>%
  as.Date(origin = "1970-01-01")

df_final <- flights_raw %>%
  filter(origin == "BWI", 
          dest %in% c("JFK", "EWR"),
          !is.na(dep_delay),
          !is.na(flight_date),
          !is.na(origin),
          !is.na(dest)) %>%
  select(
    flight_date, arr_delay, dep_delay, carrier_delay, weather_delay,
    dep_hour_local, day_of_week, month,
    dep_temp_c, dep_wind_speed_m_s, dep_ceiling_height_m,
    reporting_airline, taxi_out,
    prev_arr_delay, turnaround_minutes, aircraft_leg_number_day,
    crs_elapsed_time, actual_elapsed_time, distance
  ) %>%
  collect() %>%                        # collect before any mutates
  mutate(
    across(c(prev_arr_delay, weather_delay, carrier_delay), ~replace_na(., 0)),
    flight_date_clean = as.Date(as.character(flight_date), format = "%Y-%m-%d")
  ) %>%
  filter(!is.na(arr_delay), !is.na(dep_delay)) %>%
  mutate(
    is_delayed        = as.factor(ifelse(arr_delay >= 15, "Yes", "No")),
    is_weekend        = as.factor(ifelse(day_of_week %in% c(6, 7), "Yes", "No")),
    is_rush_hour      = as.factor(ifelse(dep_hour_local %in% c(7, 8, 16, 17, 18), "Yes", "No")),
    reporting_airline = as.factor(reporting_airline),
    buffer            = crs_elapsed_time - actual_elapsed_time,
    is_holiday        = as.factor(ifelse(as.Date(flight_date_clean) %in% holiday_windows, "Holiday", "Non-Holiday")),
    plane_size_score  = case_when(
      distance > 1500 ~ "Large",
      distance > 500  ~ "Medium",
      TRUE            ~ "Small"
    )
  )

# Output Holiday Frequency Table
holiday_freq <- as.data.frame(table(df_final$is_holiday))
names(holiday_freq) <- c("Holiday Status", "Count")

holiday_freq %>%
  knitr::kable(caption = "Flight Count by Holiday Status") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "left"
  )
```

```{r}
#| label: setup-palette
#| include: false

# ── Consistent academic colour palette ─────────────────────────
pal_2 <- c("#2166ac", "#d6604d")   # blue / coral

pal_hol <- c(
  "Non-Holiday"   = "#4d9ab5",
  "Thanksgiving"  = "#e08533",
  "Christmas"     = "#3a7d44",
  "July_4th"      = "#7b4f9e",
  "Other_Holiday" = "#b5a642"
)

theme_paper <- theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(colour = "grey40", size = 11),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )
```

### Delays by Holiday Status

Arrival delays are compared inside and outside a ±3-day window around
every 2019 US public holiday. Outliers above 100 minutes are clipped to
focus on the central distribution. The boxplot below reveals that holiday
periods produce a heavier right tail of delays, not merely a higher median
— a small share of flights incur very large delays during peak travel
windows.

```{r}
#| label: fig-holiday-boxplot
#| fig-cap: "Distribution of arrival delays for holiday versus non-holiday travel days (BWI–NYC, 2019). Whiskers extend to 1.5× IQR; outliers above 100 min are clipped."

ggplot(df_final, aes(x = is_holiday, y = arr_delay, fill = is_holiday)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, colour = "grey30") +
  coord_cartesian(ylim = c(-20, 100)) +
  scale_fill_manual(values = c("Holiday"     = pal_2[2],
                               "Non-Holiday" = pal_2[1])) +
  labs(
    title    = "The Holiday Tax: BWI–NYC Arrival Delays",
    subtitle = "±3-day windows around all 2019 US public holidays",
    x        = NULL,
    y        = "Arrival delay (minutes)"
  ) +
  theme_paper +
  theme(legend.position = "none")
```

---

## Principal Component Analysis

Before fitting non-linear models, 11 numeric predictors are standardised
(zero mean, unit variance) and reduced via PCA. Standardisation is
essential because the inputs span different units and scales — mixing
minutes, degrees Celsius, and metres per second without rescaling would
give disproportionate weight to high-magnitude variables such as
`distance`.

| Variable | Description |
|---|---|
| `dep_delay` | Departure delay (minutes) |
| `dep_hour_local` | Scheduled departure hour (local time) |
| `dep_wind_speed_m_s` | Surface wind speed at origin (m/s) |
| `dep_temp_c` | Temperature at origin (°C) |
| `dep_ceiling_height_m` | Cloud ceiling at origin (m) |
| `prev_arr_delay` | Arrival delay of the same aircraft's prior leg |
| `turnaround_minutes` | Ground time between inbound arrival and outbound departure |
| `aircraft_leg_number_day` | Leg sequence number for the aircraft that day |
| `buffer` | Scheduled elapsed time minus actual elapsed time |
| `distance` | Great-circle route distance (km) |
| `taxi_out` | Taxi-out time at origin (minutes) |

The scree plot below guides the choice to retain the first five components
for modelling, which together capture the majority of variance while
discarding noise dimensions.

```{r}
#| label: fig-pca-scree
#| fig-cap: "PCA scree plot. The elbow after component 3–4 supports retaining five components for downstream modelling."

df_pca_input <- df_final %>%
  select(dep_delay, dep_hour_local, dep_wind_speed_m_s, dep_temp_c,
         dep_ceiling_height_m, prev_arr_delay, turnaround_minutes,
         aircraft_leg_number_day, buffer, distance, taxi_out) %>%
  drop_na() %>%
  scale() %>%
  as.data.frame()

pca_result <- FactoMineR::PCA(df_pca_input, graph = FALSE)

factoextra::fviz_eig(pca_result, addlabels = TRUE,
                     barfill = pal_2[1], barcolor = "white") +
  labs(title = "PCA Scree Plot — Variance Explained per Component",
       y     = "Variance explained (%)") +
  theme_paper

pca_scores   <- as.data.frame(pca_result$ind$coord[, 1:5])
df_pca_model <- bind_cols(
  is_delayed = df_final$is_delayed[1:nrow(pca_scores)],
  pca_scores
)
```

---

## Tree-Based Modelling

Two tree-based classifiers are trained on the original feature space using
a 75/25 stratified train–test split. The target variable `is_delayed` is a
binary indicator for arrivals delayed by 15 or more minutes.

```{r}
#| label: tree-split

set.seed(42)

df_model <- df_final %>%
  select(is_delayed, dep_delay, dep_hour_local, dep_wind_speed_m_s,
         aircraft_leg_number_day, plane_size_score,
         is_holiday, is_weekend) %>%
  drop_na()

train_idx <- createDataPartition(df_model$is_delayed, p = 0.75, list = FALSE)
train_set <- df_model[ train_idx, ]
test_set  <- df_model[-train_idx, ]
```

### Decision tree

The single decision tree provides an interpretable decision path. Each
leaf node shows the predicted class and the share of training observations
reaching that node.

```{r}
#| label: fig-decision-tree
#| fig-cap: "CART decision tree. Node labels show predicted class (0 = on-time, 1 = delayed ≥15 min) and the percentage of observations at each split."
#| fig-height: 7

tree_fit <- rpart(is_delayed ~ ., data = train_set, method = "class")
rpart.plot(tree_fit, extra = 101,
           main        = "Decision Path for BWI–NYC Delays",
           box.palette = c(pal_2[1], pal_2[2]))
```

### Random forest

The random forest aggregates 100 trees and provides a more robust
variable importance estimate. `MeanDecreaseGini` measures how much each
predictor reduces node impurity across all splits, averaged over the
ensemble.

```{r}
#| label: fig-rf-varimp
#| fig-cap: "Random forest variable importance (Mean Decrease in Gini). Higher values indicate predictors that most consistently improve split purity."

rf_fit <- randomForest(is_delayed ~ ., data = train_set,
                       ntree = 100, importance = TRUE)
varImpPlot(rf_fit,
           main = "What Truly Drives Your Delays?",
           col  = pal_2[1], pch = 19, cex = 0.9)
```

---

## Non-Linear Modelling with PCA Features

Two non-linear classifiers are trained on the five PCA components derived
above. Reducing inputs to orthogonal components removes multicollinearity
and improves the stability of both SVM and neural network fits.

- **SVM (RBF kernel):** Maps PCA scores into a higher-dimensional space
  and finds a maximum-margin hyperplane. Cost parameter `C = 1` balances
  margin width against training misclassification.
- **Neural network (5 hidden neurons):** Single-hidden-layer network with
  L2 weight decay (`decay = 0.01`) to limit overfitting on the
  low-dimensional PCA input.

```{r}
#| label: tbl-nonlinear-accuracy
#| tbl-cap: "Classification accuracy of SVM and neural network models on the held-out 25% test set."

set.seed(42)
pca_idx   <- createDataPartition(df_pca_model$is_delayed, p = 0.75, list = FALSE)
train_pca <- df_pca_model[ pca_idx, ]
test_pca  <- df_pca_model[-pca_idx, ]

svm_fit   <- kernlab::ksvm(is_delayed ~ ., data = train_pca,
                           kernel = "rbfdot", C = 1)
svm_preds <- predict(svm_fit, test_pca)

nnet_fit   <- nnet::nnet(is_delayed ~ ., data = train_pca,
                         size = 5, decay = 0.01, trace = FALSE)
nnet_preds <- as.factor(predict(nnet_fit, test_pca, type = "class"))

svm_acc  <- confusionMatrix(svm_preds,  test_pca$is_delayed)$overall["Accuracy"]
nnet_acc <- confusionMatrix(nnet_preds, test_pca$is_delayed)$overall["Accuracy"]

tibble(
  Model    = c("SVM (RBF kernel)", "Neural network (5 hidden neurons)"),
  Accuracy = round(c(svm_acc, nnet_acc), 4)
) %>%
  kable(align = c("l", "r")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Holiday Category Tree Model

Holiday status is refined from a binary flag into five named categories to
give tree-based models finer-grained temporal context.

```{r}
#| label: tbl-holiday-categories
#| tbl-cap: "Distribution of flights across named holiday categories after re-coding."

df_final <- df_final %>%
  mutate(
    is_holiday = factor(case_when(
      flight_date_clean >= (as.Date("2019-11-28") - 3) &
        flight_date_clean <= (as.Date("2019-11-28") + 3) ~ "Thanksgiving",
      flight_date_clean >= (as.Date("2019-12-25") - 3) &
        flight_date_clean <= (as.Date("2019-12-25") + 3) ~ "Christmas",
      flight_date_clean >= (as.Date("2019-07-04") - 3) &
        flight_date_clean <= (as.Date("2019-07-04") + 3) ~ "July_4th",
      flight_date_clean %in% holiday_windows             ~ "Other_Holiday",
      .default                                           = "Non-Holiday"
    ), levels = c("Non-Holiday", "Thanksgiving", "Christmas",
                  "July_4th", "Other_Holiday"))
  )

table(df_final$is_holiday) %>%
  as.data.frame() %>%
  rename(`Holiday period` = Var1, `Flight count` = Freq) %>%
  arrange(desc(`Flight count`)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Hub and Airline Analysis

This section examines scheduling behaviour across the full 2019 dataset,
exploring how airlines build buffer time into published block hours and
whether that strategy differs between weekdays and weekends.

### Top destination hubs

```{r}
#| label: tbl-top-hubs
#| tbl-cap: "Top 10 destination airports by total inbound flight volume (2019)."

hub_counts <- flights_raw %>%
  select(dest, reporting_airline, taxi_in) %>%
  collect() %>%
  group_by(dest) %>%
  summarise(
    total_flights   = n(),
    unique_airlines = n_distinct(reporting_airline),
    avg_taxi_in     = mean(taxi_in, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_flights))

hub_counts %>%
  head(10) %>%
  kable(col.names = c("Destination", "Total flights",
                      "Unique airlines", "Avg taxi-in (min)"),
        digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
#| label: fig-top-hubs-all
#| fig-cap: "Top 10 destination airports by total inbound flight volume (2019)."

ggplot(head(hub_counts, 10),
       aes(x = reorder(dest, total_flights), y = total_flights)) +
  geom_col(fill = pal_2[1]) +
  coord_flip() +
  labs(title    = "Top 10 Destination Hubs — 2019",
       subtitle = "Total inbound flight arrivals",
       x = "Airport", y = "Total flights") +
  theme_paper
```

### Top destinations from BWI

```{r}
#| label: fig-top-hubs-bwi
#| fig-cap: "Top 10 destination airports from BWI by departure frequency (2019)."

hub_data <- flights_raw %>%
  filter(origin == "BWI") %>%
  select(dest) %>%
  collect() %>%
  count(dest, sort = TRUE) %>%
  slice_max(n, n = 10)

ggplot(hub_data, aes(x = reorder(dest, n), y = n)) +
  geom_col(fill = pal_2[1]) +
  coord_flip() +
  labs(title    = "Top 10 Destinations from BWI (2019)",
       subtitle = "Based on total departure frequency",
       x = "Destination", y = "Total flights") +
  theme_paper
```

### Airline scheduling honesty

The *scheduling buffer* — the difference between published block time
(`crs_elapsed_time`) and actual flight time (`actual_elapsed_time`) —
measures how much padding carriers build into their schedules. A large
buffer can absorb operational variability but may also mask chronic
performance problems.

```{r}
#| label: tbl-honesty
#| tbl-cap: "Airline scheduling buffer versus average arrival delay. High padding alongside high average delay suggests the buffer is insufficient to absorb disruption."

honesty_check <- df_final %>%
  mutate(buffer = crs_elapsed_time - actual_elapsed_time) %>%
  group_by(reporting_airline) %>%
  summarise(
    avg_padding_mins  = mean(buffer,    na.rm = TRUE),
    avg_arrival_delay = mean(arr_delay, na.rm = TRUE),
    flight_count      = n()
  ) %>%
  arrange(desc(avg_padding_mins))

honesty_check %>%
  kable(col.names = c("Airline", "Avg padding (min)",
                      "Avg arrival delay (min)", "Flights"),
        digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
#| label: fig-buffer-lollipop
#| fig-cap: "Average scheduling buffer per airline. Points to the right of zero indicate carriers that routinely pad schedules beyond actual flight time."

all_airline_buffers <- flights_raw %>%
  filter(!is.na(crs_elapsed_time), !is.na(actual_elapsed_time)) %>%
  select(reporting_airline, crs_elapsed_time, actual_elapsed_time, arr_delay) %>%
  collect() %>%
  mutate(buffer_mins = crs_elapsed_time - actual_elapsed_time) %>%
  group_by(reporting_airline) %>%
  summarise(
    avg_buffer_mins      = mean(buffer_mins, na.rm = TRUE),
    median_buffer        = median(buffer_mins, na.rm = TRUE),
    flight_count         = n(),
    on_time_arrival_rate = mean(arr_delay <= 0, na.rm = TRUE) * 100
  ) %>%
  filter(flight_count > 100) %>%
  arrange(desc(avg_buffer_mins))

ggplot(all_airline_buffers,
       aes(x = reorder(reporting_airline, avg_buffer_mins),
           y = avg_buffer_mins)) +
  geom_segment(aes(xend = reporting_airline, yend = 0), colour = "grey70") +
  geom_point(size = 4, colour = pal_2[1]) +
  coord_flip() +
  labs(title    = "Airlines: Scheduled vs Actual Block Time",
       subtitle = "Average buffer (minutes) — positive = padded schedule",
       x = "Reporting airline", y = "Average buffer (minutes)") +
  theme_paper
```

### Weekday vs weekend scheduling strategy

```{r}
#| label: fig-weekday-weekend
#| fig-cap: "Average scheduling buffer on weekdays versus weekends by airline. A larger weekend buffer may reflect anticipated congestion."

weekend_comparison <- df_final %>%
  mutate(
    buffer_mins = crs_elapsed_time - actual_elapsed_time,
    day_type    = ifelse(day_of_week %in% c(6, 7), "Weekend", "Weekday")
  ) %>%
  group_by(reporting_airline, day_type) %>%
  summarise(
    avg_buffer    = mean(buffer_mins, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay,   na.rm = TRUE),
    flight_count  = n(),
    .groups = "drop"
  )

ggplot(weekend_comparison,
       aes(x = reporting_airline, y = avg_buffer, fill = day_type)) +
  geom_col(position = "dodge", colour = "white") +
  scale_fill_manual(values = c("Weekday" = pal_2[1], "Weekend" = pal_2[2])) +
  labs(title    = "Scheduling Strategy: Weekdays vs Weekends",
       subtitle = "Average buffer (minutes) by reporting carrier",
       x = "Airline", y = "Avg buffer (minutes)", fill = NULL) +
  theme_paper
```

---

## Holiday Impact and Delay Propagation

The analysis is expanded to the full `flights_raw` dataset (all routes,
all airlines) with named holiday categories, providing a nationally
representative picture of how each holiday window affects delay rates and
where that delay originates.

```{r}
#| label: holiday-full-setup

flights_holiday <- flights_raw %>%
  filter(!is.na(arr_delay), !is.na(dep_delay)) %>%
  collect() %>%
  mutate(
    flight_date_clean = as.Date(flight_date),
    is_holiday = factor(case_when(
      flight_date_clean >= (as.Date("2019-11-28") - 3) &
        flight_date_clean <= (as.Date("2019-11-28") + 3) ~ "Thanksgiving",
      flight_date_clean >= (as.Date("2019-12-25") - 3) &
        flight_date_clean <= (as.Date("2019-12-25") + 3) ~ "Christmas",
      flight_date_clean >= (as.Date("2019-07-04") - 3) &
        flight_date_clean <= (as.Date("2019-07-04") + 3) ~ "July_4th",
      flight_date_clean %in% holiday_windows             ~ "Other_Holiday",
      .default                                           = "Non-Holiday"
    ), levels = c("Non-Holiday", "Thanksgiving", "Christmas",
                  "July_4th", "Other_Holiday"))
  )
```

### Delay summary by airline and holiday

```{r}
#| label: tbl-holiday-airline
#| tbl-cap: "Average arrival and departure delays and percentage of flights delayed ≥15 minutes, by airline and holiday period (all routes, 2019)."

holiday_airline_summary <- flights_holiday %>%
  group_by(reporting_airline, is_holiday) %>%
  summarise(
    avg_arr_delay = mean(arr_delay,       na.rm = TRUE),
    avg_dep_delay = mean(dep_delay,       na.rm = TRUE),
    pct_delayed   = mean(arr_delay >= 15, na.rm = TRUE) * 100,
    flight_count  = n(),
    .groups = "drop"
  )

holiday_airline_summary %>%
  kable(digits = 2,
        col.names = c("Airline", "Holiday period", "Avg arr delay (min)",
                      "Avg dep delay (min)", "% Delayed ≥15 min", "Flights")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(5, bold = TRUE)
```

### Delay heatmap: airline × holiday

```{r}
#| label: fig-heatmap
#| fig-cap: "Average arrival delay (minutes) by airline and holiday period. Red cells indicate worse-than-average performance; blue cells indicate better performance."
#| fig-height: 7

ggplot(holiday_airline_summary,
       aes(x = is_holiday, y = reporting_airline, fill = avg_arr_delay)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = round(avg_arr_delay, 1)),
            size = 3, colour = "white", fontface = "bold") +
  scale_fill_gradient2(low = pal_2[1], mid = "#f7f7f7", high = pal_2[2],
                       midpoint = 0, name = "Avg delay\n(min)") +
  labs(title    = "Holiday Delay Heatmap — Airline × Holiday Period",
       subtitle = "Average arrival delay (min); red = worse, blue = better",
       x = "Holiday period", y = "Airline") +
  theme_paper +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

### Share of flights delayed ≥15 minutes

```{r}
#| label: fig-pct-delayed
#| fig-cap: "Percentage of all flights delayed ≥15 minutes during each holiday window versus non-holiday days (all airlines, all routes, 2019)."

holiday_summary_overall <- flights_holiday %>%
  group_by(is_holiday) %>%
  summarise(
    avg_arr_delay = mean(arr_delay,       na.rm = TRUE),
    pct_delayed   = mean(arr_delay >= 15, na.rm = TRUE) * 100,
    flight_count  = n(),
    .groups = "drop"
  )

ggplot(holiday_summary_overall,
       aes(x = is_holiday, y = pct_delayed, fill = is_holiday)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = paste0(round(pct_delayed, 1), "%")),
            vjust = -0.5, fontface = "bold", size = 3.5) +
  scale_fill_manual(values = pal_hol) +
  labs(title    = "Share of Flights Delayed ≥15 Minutes by Holiday Window",
       subtitle = "All airlines, all routes — 2019",
       x = NULL, y = "% of flights delayed") +
  theme_paper
```

### Decomposing delay: inherited vs newly generated

A key analytical distinction is whether a flight's departure delay was
**inherited** from the previous leg flown by the same aircraft —
propagated via tail number through `prev_arr_delay` — or **freshly
generated** on the current leg. Understanding this split is important for
the predictive model: inherited delay is knowable before departure and
is captured by `prev_arr_late_15` and `rotation_continuity_flag`, whereas
newly generated delay must be modelled from operational and weather
conditions.

```{r}
#| label: propagation-setup

propagation_df <- flights_holiday %>%
  filter(!is.na(prev_arr_delay)) %>%
  mutate(
    inherited_delay = pmax(prev_arr_delay, 0),
    new_delay       = pmax(dep_delay - prev_arr_delay, 0),
    total_dep_delay = inherited_delay + new_delay
  )

propagation_summary <- propagation_df %>%
  group_by(is_holiday) %>%
  summarise(
    avg_inherited = mean(inherited_delay,  na.rm = TRUE),
    avg_new       = mean(new_delay,        na.rm = TRUE),
    avg_total     = mean(total_dep_delay,  na.rm = TRUE),
    pct_inherited = avg_inherited / avg_total * 100,
    flight_count  = n(),
    .groups = "drop"
  )
```

```{r}
#| label: tbl-propagation
#| tbl-cap: "Decomposition of average departure delay into inherited (propagated from the prior leg via tail rotation) and newly generated components, by holiday period."

propagation_summary %>%
  kable(digits = 2,
        col.names = c("Holiday period", "Avg inherited (min)",
                      "Avg new delay (min)", "Avg total dep delay (min)",
                      "% Inherited", "Flights")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(5, bold = TRUE, color = "white",
            background = spec_color(propagation_summary$pct_inherited,
                                    option = "C", direction = -1))
```

```{r}
#| label: fig-propagation-stacked
#| fig-cap: "Average departure delay decomposed into the inherited component (coral; carried from prior leg via tail rotation) and the newly generated component (blue; created on the current leg)."

propagation_summary %>%
  select(is_holiday, avg_inherited, avg_new) %>%
  pivot_longer(cols = c(avg_inherited, avg_new),
               names_to  = "delay_type",
               values_to = "minutes") %>%
  mutate(delay_type = case_match(delay_type,
    "avg_inherited" ~ "Inherited from prev. flight",
    "avg_new"       ~ "Newly generated this leg"
  )) %>%
  ggplot(aes(x = is_holiday, y = minutes, fill = delay_type)) +
  geom_col(position = "stack", colour = "white", linewidth = 0.3) +
  geom_text(aes(label = round(minutes, 1)),
            position = position_stack(vjust = 0.5),
            colour = "white", fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c(
    "Inherited from prev. flight" = pal_2[2],
    "Newly generated this leg"    = pal_2[1]
  )) +
  labs(title    = "Where Does Holiday Delay Come From?",
       subtitle = "Decomposing departure delay: inherited vs newly generated — all airlines 2019",
       x = NULL, y = "Average delay (minutes)", fill = NULL) +
  theme_paper
```

### Airlines as absorbers vs amplifiers of delay

Each airline–holiday combination is classified by whether the carrier
**absorbs** inherited disruption (points below the diagonal: newly
generated delay < inherited delay) or **amplifies** it (points above:
newly generated delay > inherited delay).

```{r}
#| label: fig-absorbers-amplifiers
#| fig-cap: "Each point represents one airline during one holiday period. Points above the dashed diagonal indicate the airline generated more new delay than it inherited, effectively amplifying disruption."
#| fig-height: 6.5

propagation_df %>%
  group_by(reporting_airline, is_holiday) %>%
  summarise(
    avg_inherited = mean(inherited_delay, na.rm = TRUE),
    avg_new       = mean(new_delay,       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = avg_inherited, y = avg_new,
             colour = is_holiday, label = reporting_airline)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey50") +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(size = 2.8, max.overlaps = 15) +
  scale_colour_manual(values = pal_hol) +
  labs(title    = "Airlines: Absorbers vs Amplifiers of Delay",
       subtitle = "Points above the diagonal → airline adds more delay than it inherited",
       x      = "Avg inherited delay (min)",
       y      = "Avg newly generated delay (min)",
       colour = "Holiday period") +
  theme_paper
```

For more details see: [github.com/zyrirena/Project](https://github.com/zyrirena/Project) 