---
title: "Example Report (R / Quarto)"
format:
  html:
    toc: true
---

## Load the dataset

```{r}
library(tidyverse)

df <- as_tibble(iris)
glimpse(df)
head(df)
```

```{r}
df %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Iris Dataset Example",
    x = "Sepal Length",
    y = "Sepal Width",
    color = "Species"
  ) +
  theme_minimal()
```

## Why this works better

Using `iris` gives you:
- no local file dependency
- no path issues during `quarto render`
- lightweight example notebooks that always work
- a familiar dataset for both Python and R

## If you want the Python notebook as Quarto-friendly too

If your Python notebook is being used inside Quarto and you want it minimal, I would use this exact version:

```python
from sklearn.datasets import load_iris
import pandas as pd

iris = load_iris(as_frame=True)
df = iris.frame.copy()
df["species"] = df["target"].map(dict(enumerate(iris.target_names)))

df.head()
```