---
title: "Flight Delay Analysis"
author: "Irena Austin"
date: "2026-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
<br>
**Part 1**
<br>
Load Libraries
<br>
"Force-load" to prevent it from crashing when running the code and used pacman to load all libraries./
Loaded parquet data from my drive.
<br>
```{r libraries}
# Force-load
if (!require("mvtnorm")) install.packages("mvtnorm")
if (!require("FactoMineR")) install.packages("FactoMineR")

library(mvtnorm)
library(FactoMineR)

# Load all other required libraries
pacman::p_load(
  tidyverse, janitor, lubridate, skimr, caret, glmnet, 
  broom, ggplot2, knitr, kableExtra, arrow, corrplot,
  rpart, rpart.plot, randomForest, ranger,
  e1071, nnet, kernlab, factoextra
)

# Load the full Parquet file
flights_raw <- open_dataset("C:/Users/ZyrIr/OneDrive/Desktop/flights_canonical_2019.parquet") %>%
  collect() %>%
  clean_names()
```
<br>
**Part 2**
<br>
Clean Data
<br>
Defined holidays (used Jonathan dates with the 3 +/- days).
Unix Epoch (1970-01-01) standard./
Filtered flights going to JFK/EWR from BWI./
Used FAA 15 minute rule to verify if flight delayed: Yes/No. Then  Looked at buffers, time of day, and tried to separate by smaller/larger planes based on distance.
<br>
```{r clean}
# Define 2019 Federal Holidays
us_holidays_2019 <- as.Date(c(
  "2019-01-01", "2019-01-21", "2019-02-18", "2019-05-27", 
  "2019-07-04", "2019-09-02", "2019-10-14", "2019-11-11", 
  "2019-11-28", "2019-12-25"
))

# Generate travel windows (+/- 3 days around each holiday)
holiday_windows <- map(us_holidays_2019, ~ seq(.x - 3, .x + 3, by = "day")) %>%
  flatten_dbl() %>%
  as.Date(origin = "1970-01-01")

df_final <- flights_raw %>%
  filter(origin == "BWI", dest %in% c("JFK", "EWR")) %>%
  select(
    flight_date, arr_delay, dep_delay, carrier_delay, weather_delay, 
    dep_hour_local, day_of_week, month, 
    dep_temp_c, dep_wind_speed_m_s, dep_ceiling_height_m,
    reporting_airline, taxi_out,
    prev_arr_delay, turnaround_minutes, aircraft_leg_number_day,
    crs_elapsed_time, actual_elapsed_time, distance
  ) %>%
  mutate(
    across(c(prev_arr_delay, weather_delay, carrier_delay), ~replace_na(., 0)),
    flight_date_clean = as.Date(flight_date)
  ) %>%
  filter(!is.na(arr_delay), !is.na(dep_delay)) %>%
  mutate(
    is_delayed   = as.factor(ifelse(arr_delay >= 15, "Yes", "No")),
    is_weekend   = as.factor(ifelse(day_of_week %in% c(6, 7), "Yes", "No")),
    is_rush_hour = as.factor(ifelse(dep_hour_local %in% c(7, 8, 16, 17, 18), "Yes", "No")),
    reporting_airline = as.factor(reporting_airline),
    buffer = crs_elapsed_time - actual_elapsed_time,
    is_holiday = as.factor(ifelse(flight_date_clean %in% holiday_windows, "Holiday", "Non-Holiday")),
    plane_size_score = case_when(
      distance > 1500 ~ "Large",
      distance > 500  ~ "Medium",
      TRUE            ~ "Small"
    )
  )

# Output Holiday Frequency Table
table(df_final$is_holiday)
```

<br>
**Part 3**
<br>
Delays by Holiday Status
<br>
Used ggplot to visualize results./
x - holiday names/
y - arrival delay in minutes/
<br>
```{r holidays}
# Visualize Arrival Delays by Holidays
ggplot(df_final, aes(x = is_holiday, y = arr_delay, fill = is_holiday)) +
  geom_boxplot(outlier.shape = NA) +
  coord_cartesian(ylim = c(-20, 100)) +
  scale_fill_manual(values = c("Holiday" = "pink", "Non-Holiday" = "darkblue")) +
  labs(title = "The Holiday Tax: BWI-NYC Arrival Delays",
       subtitle = "Comparing +/- 3 day windows for all 2019 US Holidays",
       x = "Season", y = "Arrival Delay (min)") +
  theme_minimal()
```
<br>
**Part 4**
<br>
Principal Component Analysis (PCA)
<br>
scaling to turn each value into a score to compare all together; helps to reduce the noise of all the data.
<br>
```{r PCA}
# Standardize numeric inputs
df_pca_input <- df_final %>%
  select(dep_delay, dep_hour_local, dep_wind_speed_m_s, dep_temp_c,
         dep_ceiling_height_m, prev_arr_delay, turnaround_minutes,
         aircraft_leg_number_day, buffer, distance, taxi_out) %>%
  drop_na() %>%
  scale() %>% 
  as.data.frame()

# Execute PCA via FactoMineR
pca_result <- FactoMineR::PCA(df_pca_input, graph = FALSE)

# Scree Plot
factoextra::fviz_eig(pca_result, addlabels = TRUE) + 
  labs(title = "PCA Scree Plot", y = "% Variance")

# Extract top 5 components for nonlinear modeling
pca_scores <- as.data.frame(pca_result$ind$coord[, 1:5])
df_pca_model <- bind_cols(is_delayed = df_final$is_delayed[1:nrow(pca_scores)], pca_scores)
```
<br>
**Part 5**
<br>
Tree Modeling
<br>


<br>
```{r tree-based modeling}
set.seed(42)

# Select features for the original coordinate system
df_model <- df_final %>%
  select(is_delayed, dep_delay, dep_hour_local, dep_wind_speed_m_s, 
         aircraft_leg_number_day, plane_size_score, is_holiday, is_weekend) %>%
  drop_na()

train_idx <- createDataPartition(df_model$is_delayed, p = 0.75, list = FALSE)
train_set <- df_model[train_idx, ]
test_set  <- df_model[-train_idx, ]

# Decision Tree
tree_fit <- rpart(is_delayed ~ ., data = train_set, method = "class")
rpart.plot(tree_fit, extra = 101, main = "Decision Path for BWI-NYC Delays")

# Random Forest
rf_fit <- randomForest(is_delayed ~ ., data = train_set, ntree = 100, importance = TRUE)
varImpPlot(rf_fit, main = "What Truly Drives Your Delays?")
```
<br>
**Part 6**
<br>
Non Linear Modeling
<br>


<br>
```{r nonlinear modeling with PCA features}
# Split PCA data
pca_train_idx <- createDataPartition(df_pca_model$is_delayed, p = 0.75, list = FALSE)
train_pca <- df_pca_model[pca_train_idx, ]
test_pca  <- df_pca_model[-pca_train_idx, ]

# SVM (Radial Basis Function Kernel)
svm_fit <- kernlab::ksvm(is_delayed ~ ., data = train_pca, kernel = "rbfdot", C = 1)
svm_preds <- predict(svm_fit, test_pca)

# Neural Network (5 Hidden Neurons)
nnet_fit <- nnet::nnet(is_delayed ~ ., data = train_pca, size = 5, decay = 0.01, trace = FALSE)
nnet_preds <- as.factor(predict(nnet_fit, test_pca, type = "class"))

# Accuracy Comparison
cat("SVM Accuracy:", confusionMatrix(svm_preds, test_pca$is_delayed)$overall["Accuracy"], "\n")
cat("Neural Net Accuracy:", confusionMatrix(nnet_preds, test_pca$is_delayed)$overall["Accuracy"], "\n")
```

<br>
**Part 7**
<br>
Using Tree modeling with holidays
<br>


<br>.

```{r tree model for holidays}
# Update is_holiday in df_final with named holiday categories
df_final <- df_final %>%
  mutate(
    is_holiday = as.factor(case_when(
      flight_date_clean >= (as.Date("2019-11-28") - 3) & 
      flight_date_clean <= (as.Date("2019-11-28") + 3) ~ "Thanksgiving",
      
      flight_date_clean >= (as.Date("2019-12-25") - 3) & 
      flight_date_clean <= (as.Date("2019-12-25") + 3) ~ "Christmas",
      
      flight_date_clean >= (as.Date("2019-07-04") - 3) & 
      flight_date_clean <= (as.Date("2019-07-04") + 3) ~ "July4th",
      
      flight_date_clean %in% holiday_windows ~ "Other_Holiday",
      
      TRUE ~ "Non-Holiday"
    ))
  )

# Verify the distribution
table(df_final$is_holiday)
```
<br>
**Part 8**
<br>
Filtering 
<br>
Filter the data, looking at all large hubs by destination and flight number./


<br>
```{r identify big hubs}
# Filter for BWI flights
df_clean <- flights_raw %>% 
  filter(origin == "BWI", dest %in% c("JFK", "EWR"))

# Count flights per destination
hub_counts <- flights_raw %>%
  group_by(dest) %>%
  summarise(
    total_flights = n(),
    unique_airlines = n_distinct(reporting_airline),
    avg_taxi_in = mean(taxi_in, na.rm = TRUE)
  ) %>%
  arrange(desc(total_flights))

head(hub_counts, 10)

# Top 10 Hubs Overall
ggplot(head(hub_counts, 10), aes(x = reorder(dest, total_flights), y = total_flights, fill = total_flights)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "lightpink", high = "darkred") +
  labs(
    title = "Top 10 Destination Hubs in the 2019 Dataset",
    subtitle = "Analysis of total flight arrivals by airport code",
    x = "Airport Code",
    y = "Total Number of Flights"
  ) +
  theme_minimal()

# Top 10 Hubs from BWI
hub_data <- flights_raw %>%
  filter(origin == "BWI") %>%
  count(dest, sort = TRUE) %>%
  top_n(10, n)

ggplot(hub_data, aes(x = reorder(dest, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "lightpink", high = "darkred") +
  labs(
    title = "Top 10 Destination Hubs from BWI (2019)",
    subtitle = "Analysis based on total flight frequency",
    x = "Airport Code (Destination)",
    y = "Total Number of Flights"
  ) +
  theme_minimal()

# Airline Scheduling Honesty Check
honesty_check <- df_final %>%
  mutate(buffer = crs_elapsed_time - actual_elapsed_time) %>%
  group_by(reporting_airline) %>%
  summarise(
    avg_padding_mins = mean(buffer, na.rm = TRUE),
    avg_arrival_delay = mean(arr_delay, na.rm = TRUE),
    flight_count = n()
  ) %>%
  arrange(desc(avg_padding_mins))

kable(honesty_check, caption = "Airline Scheduling Honesty (Padding vs. Delay)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Airline Market Share
airline_counts <- flights_raw %>%
  group_by(reporting_airline) %>%
  summarise(total_flights = n()) %>%
  arrange(desc(total_flights))

head(airline_counts)

ggplot(head(airline_counts, 10), aes(x = reorder(reporting_airline, total_flights), y = total_flights, fill = total_flights)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "lightpink", high = "darkred") +
  labs(
    title = "Airline Market Share (2019 Dataset)",
    subtitle = "Total flight volume by reporting carrier",
    x = "Airline Code",
    y = "Total Number of Flights"
  ) +
  theme_minimal()

# Target Airline Buffer Analysis
target_airlines <- c("WN", "DL", "AA", "OO", "UA", "YX", "MQ", "B6", "OH", "AS")

airline_buffer_analysis <- df_final %>%
  filter(reporting_airline %in% target_airlines) %>%
  mutate(buffer_mins = crs_elapsed_time - actual_elapsed_time) %>%
  group_by(reporting_airline) %>%
  summarise(
    avg_padding = mean(buffer_mins, na.rm = TRUE),
    median_padding = median(buffer_mins, na.rm = TRUE),
    on_time_rate = mean(arr_delay <= 0, na.rm = TRUE) * 100,
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    flight_count = n()
  ) %>%
  arrange(desc(avg_padding))

airline_buffer_analysis %>%
  knitr::kable(caption = "Airline Scheduling Buffer Analysis (2019)",
               digits = 2,
               col.names = c("Airline", "Avg Padding (min)", "Median Padding", "On-Time %", "Avg Delay", "Flights"))

# All Airline Buffer Analysis
all_airline_buffers <- flights_raw %>%
  filter(!is.na(crs_elapsed_time), !is.na(actual_elapsed_time)) %>%
  mutate(buffer_mins = crs_elapsed_time - actual_elapsed_time) %>%
  group_by(reporting_airline) %>%
  summarise(
    avg_buffer_mins = mean(buffer_mins, na.rm = TRUE),
    median_buffer   = median(buffer_mins, na.rm = TRUE),
    flight_count    = n(),
    on_time_arrival_rate = mean(arr_delay <= 0, na.rm = TRUE) * 100
  ) %>%
  filter(flight_count > 100) %>%
  arrange(desc(avg_buffer_mins))

print(all_airline_buffers)

ggplot(all_airline_buffers, aes(x = reorder(reporting_airline, avg_buffer_mins), y = avg_buffer_mins)) +
  geom_segment(aes(xend = reporting_airline, yend = 0), color = "grey") +
  geom_point(size = 4, aes(color = avg_buffer_mins)) +
  scale_color_gradient(low = "lightpink", high = "purple") +
  coord_flip() +
  labs(
    title = "Which Airlines Build the Most 'Rest Time' into their Schedules?",
    subtitle = "Average Buffer (Scheduled Time minus Actual Flight Time)",
    x = "Reporting Airline",
    y = "Average Buffer (Minutes)"
  ) +
  theme_minimal()

# Weekday vs Weekend Buffer Comparison
weekend_comparison <- df_final %>%
  mutate(
    buffer_mins = crs_elapsed_time - actual_elapsed_time,
    day_type = ifelse(day_of_week %in% c(6, 7), "Weekend", "Weekday")
  ) %>%
  group_by(reporting_airline, day_type) %>%
  summarise(
    avg_buffer = mean(buffer_mins, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    flight_count = n()
  ) %>%
  arrange(reporting_airline, day_type)

print(weekend_comparison)

ggplot(weekend_comparison, aes(x = reporting_airline, y = avg_buffer, fill = day_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "steelblue", "Weekend" = "orange")) +
  labs(
    title = "Airline Scheduling Strategy: Weekdays vs. Weekends",
    subtitle = "Average Buffer (Minutes) by Reporting Carrier",
    x = "Airline Code",
    y = "Avg Buffer (Minutes)"
  ) +
  theme_minimal()
```
<br>
**Part 9**
<br>
Holiday Impact
<br>
Looked at the whole dataset again.
<br>
```{r holiday impact on delays}
# re-run the named holiday tagging on the full flights_raw dataset
# all routes
flights_holiday <- flights_raw %>%
  filter(!is.na(arr_delay), !is.na(dep_delay)) %>%
  mutate(
    flight_date_clean = as.Date(flight_date),
    is_holiday = as.factor(case_when(
      flight_date_clean >= (as.Date("2019-11-28") - 3) &
      flight_date_clean <= (as.Date("2019-11-28") + 3) ~ "Thanksgiving",

      flight_date_clean >= (as.Date("2019-12-25") - 3) &
      flight_date_clean <= (as.Date("2019-12-25") + 3) ~ "Christmas",

      flight_date_clean >= (as.Date("2019-07-04") - 3) &
      flight_date_clean <= (as.Date("2019-07-04") + 3) ~ "July_4th",

      flight_date_clean %in% holiday_windows ~ "Other_Holiday",
      TRUE ~ "Non-Holiday"
    )),
    # Reorder factor so Non-Holiday is the baseline reference
    is_holiday = factor(is_holiday, levels = c(
      "Non-Holiday", "Thanksgiving", "Christmas", "July_4th", "Other_Holiday"
    ))
  )

# average delays by holiday
holiday_airline_summary <- flights_holiday %>%
  group_by(reporting_airline, is_holiday) %>%
  summarise(
    avg_arr_delay  = mean(arr_delay,  na.rm = TRUE),
    avg_dep_delay  = mean(dep_delay,  na.rm = TRUE),
    pct_delayed    = mean(arr_delay >= 15, na.rm = TRUE) * 100,
    flight_count   = n(),
    .groups = "drop"
  )

# Print the full summary
kable(holiday_airline_summary,
      caption = "Holiday Impact on Delays by Airline (All Routes, 2019)",
      digits = 2,
      col.names = c("Airline", "Holiday Period", "Avg Arr Delay",
                    "Avg Dep Delay", "% Flights Delayed (≥15 min)", "Flights")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(5, bold = TRUE)

# Heatmap of Avg Arrival Delay by Airline × Holiday
ggplot(holiday_airline_summary,
       aes(x = is_holiday, y = reporting_airline, fill = avg_arr_delay)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(avg_arr_delay, 1)), size = 3, color = "white") +
  scale_fill_gradient2(
    low = "steelblue", mid = "lightyellow", high = "firebrick",
    midpoint = 0, name = "Avg Delay\n(min)"
  ) +
  labs(
    title = "Holiday Delay Heatmap: Every Airline vs Every Holiday",
    subtitle = "Average arrival delay (minutes) — red = worse, blue = better",
    x = "Holiday Period", y = "Airline"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))

# Flights Delayed ≥15 min by Holiday Period 
holiday_summary_overall <- flights_holiday %>%
  group_by(is_holiday) %>%
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    pct_delayed   = mean(arr_delay >= 15, na.rm = TRUE) * 100,
    flight_count  = n(),
    .groups = "drop"
  )

ggplot(holiday_summary_overall,
       aes(x = is_holiday, y = pct_delayed, fill = is_holiday)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct_delayed, 1), "%")),
            vjust = -0.5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Non-Holiday"  = "pink",
    "Thanksgiving" = "darkorange",
    "Christmas"    = "green",
    "July4th"      = "blue",
    "Other_Holiday"= "yellow"
  )) +
  labs(
    title = "Share of Flights Delayed ≥15 Minutes by Holiday Window",
    subtitle = "All airlines, all routes — 2019",
    x = "Holiday Period", y = "% of Flights Delayed"
  ) +
  theme_minimal()


# propagated delays, how much of this flight's arrival delay was inherited
# using tail number (prev_arr_delay),
# versus new delay generated on this leg (dep_delay - prev_arr_delay).

propagation_df <- flights_holiday %>%
  filter(!is.na(prev_arr_delay)) %>%
  mutate(
    # Inherited delay: how much was carried in from previous leg
    inherited_delay = pmax(prev_arr_delay, 0),

    # Freshly generated delay: new delay created on THIS leg
    new_delay = pmax(dep_delay - prev_arr_delay, 0),

    # Total departure delay decomposed
    total_dep_delay = inherited_delay + new_delay
  )

# Inherited vs New Delay by Holiday 
propagation_summary <- propagation_df %>%
  group_by(is_holiday) %>%
  summarise(
    avg_inherited_delay = mean(inherited_delay, na.rm = TRUE),
    avg_new_delay       = mean(new_delay,       na.rm = TRUE),
    avg_total_dep_delay = mean(total_dep_delay, na.rm = TRUE),
    pct_inherited       = mean(inherited_delay, na.rm = TRUE) /
                          mean(total_dep_delay, na.rm = TRUE) * 100,
    flight_count        = n(),
    .groups = "drop"
  )

kable(propagation_summary,
      caption = "Propagated vs. Freshly Generated Delays by Holiday Period",
      digits = 2,
      col.names = c("Holiday Period", "Avg Inherited (min)",
                    "Avg New Delay (min)", "Avg Total Dep Delay",
                    "% That Is Inherited", "Flights")) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(5, bold = TRUE, color = "white",
              background = spec_color(propagation_summary$pct_inherited,
                                      option = "C", direction = -1))

# Stacked Bar — Inherited vs New Delay by Holiday
propagation_long <- propagation_summary %>%
  select(is_holiday, avg_inherited_delay, avg_new_delay) %>%
  pivot_longer(cols = c(avg_inherited_delay, avg_new_delay),
               names_to  = "delay_type",
               values_to = "minutes") %>%
  mutate(delay_type = recode(delay_type,
    "avg_inherited_delay" = "Inherited from Prev. Flight",
    "avg_new_delay"       = "Newly Generated This Leg"
  ))

ggplot(propagation_long,
       aes(x = is_holiday, y = minutes, fill = delay_type)) +
  geom_col(position = "stack") +
  geom_text(aes(label = round(minutes, 1)),
            position = position_stack(vjust = 0.5),
            color = "white", fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c(
    "Inherited from Prev. Flight" = "firebrick",
    "Newly Generated This Leg"    = "steelblue"
  )) +
  labs(
    title = "Where Does Holiday Delay Come From?",
    subtitle = "Decomposing departure delay into inherited vs. newly generated — all airlines 2019",
    x = "Holiday Period", y = "Average Delay (Minutes)",
    fill = "Delay Source"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# propagated delay by airline
propagation_airline <- propagation_df %>%
  group_by(reporting_airline, is_holiday) %>%
  summarise(
    avg_inherited = mean(inherited_delay, na.rm = TRUE),
    avg_new       = mean(new_delay,       na.rm = TRUE),
    .groups = "drop"
  )

ggplot(propagation_airline,
       aes(x = avg_inherited, y = avg_new,
           color = is_holiday, label = reporting_airline)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text(vjust = -0.7, size = 2.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              color = "grey40") +
  scale_color_manual(values = c(
    "Non-Holiday"   = "pink",
    "Thanksgiving"  = "darkorange",
    "Christmas"     = "green",
    "July4th"       = "blue",
    "Other_Holiday" = "yellow"
  )) +
  labs(
    title = "Airlines: Absorbers vs. Amplifiers of Delay",
    subtitle = "Points above the diagonal = airline adds MORE delay than it inherited",
    x = "Avg Inherited Delay (min)",
    y = "Avg Newly Generated Delay (min)",
    color = "Holiday Period"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```