---
title: "OR568_Final_2026_0331"
author: "V Sokolow"
date: "2026-03-31"
output:
  word_document: default
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
pacman::p_load(dplyr, skimr, tidyverse, caret, glmnet, janitor, ggplot2, lubridate, purrr,broom, arrow, recipes)

#Select columns to keep
bts_info <- c("year","month","dayof_month","day_of_week","flight_date","origin","dest","cancelled","dep_time_blk","crs_dep_time", "dep_time","dep_delay","taxi_out","wheels_off","air_time","wheels_on","arr_time_blk","crs_arr_time", "arr_time","arr_delay","taxi_in","distance","reporting_airline","dot_id_reporting_airline","iata_code_reporting_airline","tail_number")

arr_weather_info <- c("arr_temp_c","arr_wind_speed_m_s","arr_wind_dir_deg")
dep_weather_info <- c("dep_temp_c","dep_wind_speed_m_s","dep_wind_dir_deg")

enriched_flights_NEW_2019 <- read_parquet("flights_canonical_2019.parquet") %>% 
  clean_names() %>%
  select(bts_info, arr_weather_info, dep_weather_info)
```

```{r}
US_HOLIDAYS_2019 <- as_date(c(
  "2019-01-01",  # New Year's Day
  "2019-01-21",  # MLK Day
  "2019-02-18",  # Presidents Day
  "2019-05-27",  # Memorial Day
  "2019-07-04",  # Independence Day
  "2019-09-02",  # Labor Day
  "2019-10-14",  # Columbus Day
  "2019-11-11",  # Veterans Day
  "2019-11-28",  # Thanksgiving
  "2019-12-25"   # Christmas
))

# Create ±3‑day windows for each holiday
holiday_windows <- map(US_HOLIDAYS_2019, ~ (.x - 3):(.x + 3)) %>%
  flatten_int() %>%
  as_date()


flights_filtered <- enriched_flights_NEW_2019 %>%
  # create flag for previous flight delayed
  mutate(
    dep_ts = ymd_hm(paste(flight_date, sprintf("%04d", dep_time))),
    arr_ts = ymd_hm(paste(flight_date, sprintf("%04d", arr_time))),
    # Fix overnight arrivals (arr_time < dep_time)
    arr_ts = if_else(arr_ts < dep_ts, arr_ts + days(1), arr_ts),
    # Flag holidays
    is_holiday = as.integer(flight_date %in% holiday_windows)
  ) %>%
  arrange(tail_number, dep_ts) %>%
  group_by(tail_number) %>%
  mutate(
    prev_arr_delay = lag(arr_delay),
    prev_arr_ts    = lag(arr_ts),
    time_since_prev = as.numeric(difftime(dep_ts, prev_arr_ts, units = "hours")),
    prev_flight_late = prev_arr_delay > 15 & time_since_prev <= 8
  ) %>%
  ungroup() %>%
  mutate(
    is_dep_delay = factor(dep_delay > 15, levels = c(FALSE, TRUE)),
    is_arr_delay = factor(arr_delay > 15, levels = c(FALSE, TRUE)),
    hour_of_day = dep_time %/% 100,
    # Fallback to lubridate if day_of_week column is absent
    is_weekend = day_of_week %in% c("6", "7"),
    ) %>%
  group_by(hour_of_day) %>%
  mutate(congestion = n()) %>%
  ungroup %>%
  filter(cancelled == '0')



# Define the 5-fold Cross-Validation strategy
set.seed(42)
train_control <- trainControl(method = "cv", number = 5)
```

## Models
### Departure Delay

```{r}
# Limit data to relevant variables
dep_flight_filtered <- flights_filtered %>%
  filter(cancelled == '0') %>%
  filter(origin == "BWI") %>%
  select(c("dep_delay","is_dep_delay","prev_flight_late","dep_temp_c","dep_wind_speed_m_s","reporting_airline","dep_time_blk","is_weekend","is_holiday")) %>%
  # Scale weather vars
  mutate(across(all_of(c("dep_temp_c","dep_wind_speed_m_s")), ~ as.numeric(scale(.))))


dep_flight <- dep_flight_filtered %>%
  filter(
    !is.na(dep_delay),
    !is.na(prev_flight_late),    
    dep_delay >= -60,
    dep_delay <= 1440,
    !is.na(dep_temp_c),
    !is.na(dep_wind_speed_m_s)

  )

```

#### Regression
##### Linear

```{r, warning = FALSE}
model_lm <- train(
  dep_delay ~ prev_flight_late + dep_temp_c + dep_wind_speed_m_s + reporting_airline + dep_time_blk + is_weekend + is_holiday,
  data = dep_flight, method = "lm", trControl = train_control
)

interpret_lm <- function(model) {
  cat("\n=== Linear Model Summary ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)
}
interpret_lm(model_lm)
```

##### Linear + PCA

```{r}
rec <- recipe(
  dep_delay ~ prev_flight_late + dep_temp_c + dep_wind_speed_m_s +
    reporting_airline + dep_time_blk + is_weekend + is_holiday,
  data = dep_flight
) %>%
  step_normalize(dep_temp_c, dep_wind_speed_m_s) %>%
  step_pca(dep_temp_c, dep_wind_speed_m_s, num_comp = 2)

model_lm <- train(
  rec,
  data = dep_flight,
  method = "lm",
  trControl = train_control
)

interpret_pca_lm <- function(model) {
  cat("\n=== Linear Model Summary (with PCA) ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)

  cat("\n=== PCA Loadings ===\n")
  print(tidy(model$recipe, number = 2))   # adjust step number if needed

  cat("\n=== PCA Variance Explained ===\n")
  print(model$recipe$steps[[2]]$res)      # variance explained
}


interpret_pca_lm(model_lm)
```


#### Classification
##### Logistic Regression

```{r}
model_logit <- train(
  is_dep_delay ~ prev_flight_late + dep_temp_c + dep_wind_speed_m_s + reporting_airline + dep_time_blk + is_weekend + is_holiday,
  data = dep_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)

interpret_classification <- function(model, data) {
  cat("\n=== Classification Performance ===\n")
  print(model$results)

  cat("\n=== Confusion Matrix ===\n")
  preds <- predict(model, newdata = dep_flight)
  print(caret::confusionMatrix(preds, dep_flight$is_dep_delay))

  if ("prob" %in% model$control$classProbs) {
    cat("\n=== ROC / AUC ===\n")
    probs <- predict(model, newdata = data, type = "prob")[,2]
    roc_obj <- pROC::roc(response = data$delay_flag, predictor = probs)
    print(pROC::auc(roc_obj))
  }
}

interpret_classification(model_logit)

```

```{r}
# Predicted vs Actual — Linear Regression
dep_flight$predictions <- predict(model_lm, newdata = dep_flight)

ggplot(dep_flight, aes(x = dep_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Departure Delay",
    x = "Actual Departure Delay (min)",
    y = "Predicted Departure Delay (min)"
  ) +
  theme_minimal()
```



### Arrival Delay

```{r}
# Limit data to relevant variables
arr_flight_filtered <- flights_filtered %>%
  filter(cancelled == '0') %>%
  filter(dest == "BWI") %>%
  select(c("arr_delay","is_arr_delay","is_dep_delay","arr_temp_c","arr_wind_speed_m_s","reporting_airline","arr_time_blk","is_weekend","congestion","is_holiday")) %>%
  # Scale weather vars
  mutate(across(all_of(c("arr_temp_c","arr_wind_speed_m_s")), ~ as.numeric(scale(.))))


arr_flight <- arr_flight_filtered %>%
  filter(
    !is.na(arr_delay),
    !is.na(is_dep_delay),    
    arr_delay >= -60,
    arr_delay <= 1440,
    !is.na(arr_temp_c),
    !is.na(arr_wind_speed_m_s)

  )

```

#### Regression
##### Linear

```{r}
model_lm <- train(
  arr_delay ~ is_dep_delay + arr_temp_c + arr_wind_speed_m_s + reporting_airline + arr_time_blk + is_weekend + congestion + is_holiday,
  data = arr_flight, method = "lm", trControl = train_control
)

interpret_lm <- function(model) {
  cat("\n=== Linear Model Summary ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)
}
interpret_lm(model_lm)
```

##### Linear + PCA

```{r}
rec <- recipe(
  arr_delay ~ is_dep_delay + arr_temp_c + arr_wind_speed_m_s + reporting_airline + arr_time_blk + is_weekend + congestion + is_holiday,
  data = arr_flight
) %>%
  step_normalize(arr_temp_c, arr_wind_speed_m_s) %>%
  step_pca(arr_temp_c, arr_wind_speed_m_s, num_comp = 2)

model_lm <- train(
  rec,
  data = arr_flight,
  method = "lm",
  trControl = train_control
)


interpret_pca_lm <- function(model) {
  cat("\n=== Linear Model Summary (with PCA) ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)

  cat("\n=== PCA Loadings ===\n")
  print(tidy(model$recipe, number = 2))   # adjust step number if needed

  cat("\n=== PCA Variance Explained ===\n")
  print(model$recipe$steps[[2]]$res)      # variance explained
}


interpret_pca_lm(model_lm)
```


#### Classification
##### Logistic Regression

```{r}
model_logit <- train(
  is_arr_delay ~ is_dep_delay + arr_temp_c + arr_wind_speed_m_s + reporting_airline + arr_time_blk + is_weekend + congestion + is_holiday,
  data = arr_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)

interpret_classification <- function(model, data) {
  cat("\n=== Classification Performance ===\n")
  print(model$results)

  cat("\n=== Confusion Matrix ===\n")
  preds <- predict(model, newdata = arr_flight)
  print(caret::confusionMatrix(preds, arr_flight$is_arr_delay))

  if ("prob" %in% model$control$classProbs) {
    cat("\n=== ROC / AUC ===\n")
    probs <- predict(model, newdata = data, type = "prob")[,2]
    roc_obj <- pROC::roc(response = data$delay_flag, predictor = probs)
    print(pROC::auc(roc_obj))
  }
}

interpret_classification(model_logit)
```


##### Arrival Delay Performance 

```{r}
# Predicted vs Actual — Linear Regression
arr_flight$predictions <- predict(model_lm, newdata = arr_flight)

ggplot(arr_flight, aes(x = arr_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Arrival Delay",
    x = "Actual Arrival Delay (min)",
    y = "Predicted Arrival Delay (min)"
  ) +
  theme_minimal()
```

#### Non-inherited delays

##### Departure Delay where previous incoming flight is not delayed
```{r}
dep_flight <- dep_flight %>%
  filter(prev_flight_late == FALSE)
```


#### Regression
##### Linear

```{r, warning = FALSE}
model_lm <- train(
  dep_delay ~ dep_temp_c + dep_wind_speed_m_s + reporting_airline + dep_time_blk + is_weekend + is_holiday,
  data = dep_flight, method = "lm", trControl = train_control
)

interpret_lm <- function(model) {
  cat("\n=== Linear Model Summary ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)
}
interpret_lm(model_lm)
```

##### Linear + PCA

```{r}
rec <- recipe(
  dep_delay ~ dep_temp_c + dep_wind_speed_m_s +
    reporting_airline + dep_time_blk + is_weekend + is_holiday,
  data = dep_flight
) %>%
  step_normalize(dep_temp_c, dep_wind_speed_m_s) %>%
  step_pca(dep_temp_c, dep_wind_speed_m_s, num_comp = 2)

model_lm <- train(
  rec,
  data = dep_flight,
  method = "lm",
  trControl = train_control
)

interpret_pca_lm <- function(model) {
  cat("\n=== Linear Model Summary (with PCA) ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)

  cat("\n=== PCA Loadings ===\n")
  print(tidy(model$recipe, number = 2))   # adjust step number if needed

  cat("\n=== PCA Variance Explained ===\n")
  print(model$recipe$steps[[2]]$res)      # variance explained
}


interpret_pca_lm(model_lm)
```


#### Classification
##### Logistic Regression

```{r}
model_logit <- train(
  is_dep_delay ~ dep_temp_c + dep_wind_speed_m_s + reporting_airline + dep_time_blk + is_weekend + is_holiday,
  data = dep_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)

interpret_classification <- function(model, data) {
  cat("\n=== Classification Performance ===\n")
  print(model$results)

  cat("\n=== Confusion Matrix ===\n")
  preds <- predict(model, newdata = dep_flight)
  print(caret::confusionMatrix(preds, dep_flight$is_dep_delay))

  if ("prob" %in% model$control$classProbs) {
    cat("\n=== ROC / AUC ===\n")
    probs <- predict(model, newdata = data, type = "prob")[,2]
    roc_obj <- pROC::roc(response = data$delay_flag, predictor = probs)
    print(pROC::auc(roc_obj))
  }
}

interpret_classification(model_logit)

```

```{r}
# Predicted vs Actual — Linear Regression
dep_flight$predictions <- predict(model_lm, newdata = dep_flight)

ggplot(dep_flight, aes(x = dep_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Departure Delay",
    x = "Actual Departure Delay (min)",
    y = "Predicted Departure Delay (min)"
  ) +
  theme_minimal()
```

##### Arrival Delay where flight not delayed upon departure
```{r}
arr_flight <- arr_flight %>%
  filter(is_dep_delay == FALSE)
```


#### Regression
##### Linear

```{r}
model_lm <- train(
  arr_delay ~ arr_temp_c + arr_wind_speed_m_s + reporting_airline + arr_time_blk + is_weekend + congestion + is_holiday,
  data = arr_flight, method = "lm", trControl = train_control
)

interpret_lm <- function(model) {
  cat("\n=== Linear Model Summary ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)
}
interpret_lm(model_lm)
```

##### Linear + PCA

```{r}
rec <- recipe(
  arr_delay ~ arr_temp_c + arr_wind_speed_m_s + reporting_airline + arr_time_blk + is_weekend + congestion + is_holiday,
  data = arr_flight
) %>%
  step_normalize(arr_temp_c, arr_wind_speed_m_s) %>%
  step_pca(arr_temp_c, arr_wind_speed_m_s, num_comp = 2)

model_lm <- train(
  rec,
  data = arr_flight,
  method = "lm",
  trControl = train_control
)


interpret_pca_lm <- function(model) {
  cat("\n=== Linear Model Summary (with PCA) ===\n")
  print(summary(model$finalModel))

  cat("\n=== Coefficients (sorted by |estimate|) ===\n")
  coef_tbl <- broom::tidy(model$finalModel) %>%
    dplyr::mutate(abs_est = abs(estimate)) %>%
    dplyr::arrange(desc(abs_est))
  print(coef_tbl)

  cat("\n=== PCA Loadings ===\n")
  print(tidy(model$recipe, number = 2))   # adjust step number if needed

  cat("\n=== PCA Variance Explained ===\n")
  print(model$recipe$steps[[2]]$res)      # variance explained
}


interpret_pca_lm(model_lm)
```


#### Classification
##### Logistic Regression

```{r}
model_logit <- train(
  is_arr_delay ~ arr_temp_c + arr_wind_speed_m_s + reporting_airline + arr_time_blk + is_weekend + congestion + is_holiday,
  data = arr_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)

interpret_classification <- function(model, data) {
  cat("\n=== Classification Performance ===\n")
  print(model$results)

  cat("\n=== Confusion Matrix ===\n")
  preds <- predict(model, newdata = arr_flight)
  print(caret::confusionMatrix(preds, arr_flight$is_arr_delay))

  if ("prob" %in% model$control$classProbs) {
    cat("\n=== ROC / AUC ===\n")
    probs <- predict(model, newdata = data, type = "prob")[,2]
    roc_obj <- pROC::roc(response = data$delay_flag, predictor = probs)
    print(pROC::auc(roc_obj))
  }
}

interpret_classification(model_logit)
```


##### Arrival Delay Performance 

```{r}
# Predicted vs Actual — Linear Regression
arr_flight$predictions <- predict(model_lm, newdata = arr_flight)

ggplot(arr_flight, aes(x = arr_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Arrival Delay",
    x = "Actual Arrival Delay (min)",
    y = "Predicted Arrival Delay (min)"
  ) +
  theme_minimal()
```

