---
title: "OR568_Midterm_2026_0310"
author: "V Sokolow"
date: "2026-03-10"
format:
  html: default
  docx: default
execute:
  eval: false
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup/Load Data

```{r}
pacman::p_load(dplyr, skimr, tidyverse, caret, glmnet, janitor, ggplot2, lubridate, purrr,broom)

# enriched_flights_NEW_2019 <- read.csv("C:\\Users\\valso\\OneDrive - George Mason University - O365 Production\\2026_T1\\GitHub Clone\\OR568_ML_Project\\shared-notebooks\\notebooks\\val\\r\\enriched_flights_NEW.csv", 
#                                       stringsAsFactors = FALSE)

###################################################
# Code to Load Data from S3
###################################################
library(dplyr)
library(janitor)
notebook_dir <- dirname(knitr::current_input(dir = TRUE))

util_path <- normalizePath(
  file.path(
    notebook_dir,
    "..", "..", "..",
    "common_utils", "r", "load_flight_data.r"
  ),
  winslash = "/",
  mustWork = TRUE
)

source(util_path)

# Load data and clean column names
enriched_flights_NEW_2019 <- load_flight_data() %>%
  janitor::clean_names()

###################################################
# End Code to Load Data from S3
###################################################
```

## Data Prep

```{r}
#| eval: false
#Assign imported dataset
flights_raw <- enriched_flights_NEW_2019 %>%
  clean_names()

#Select columns to keep
bts_info <- c("year","month","dayof_month","day_of_week","flight_date","origin","dest","cancelled","dep_time_blk","crs_dep_time", "dep_time","dep_delay","taxi_out","wheels_off","air_time","wheels_on","arr_time_blk","crs_arr_time", "arr_time","arr_delay","taxi_in","distance","reporting_airline","dot_id_reporting_airline","iata_code_reporting_airline","tail_number")

arr_weather_info <- c("arr_tmpf","arr_dwpf","arr_relh","arr_sknt","arr_vsby")
dep_weather_info <- c("dep_tmpf","dep_dwpf","dep_relh","dep_sknt","dep_vsby")

plane_info <- c("aircraft_manufacturer","aircraft_model","aircraft_type","weight_category","num_engines","engine_type")

flights_filtered <- flights_raw %>%
  select(bts_info, arr_weather_info, dep_weather_info, plane_info) %>%
  # create flag for previous flight delayed
  mutate(
    dep_ts = ymd_hm(paste(flight_date, sprintf("%04d", dep_time))),
    arr_ts = ymd_hm(paste(flight_date, sprintf("%04d", arr_time))),
    # Fix overnight arrivals (arr_time < dep_time)
    arr_ts = if_else(arr_ts < dep_ts, arr_ts + days(1), arr_ts)
  ) %>%
  arrange(tail_number, dep_ts) %>%
  group_by(tail_number) %>%
  mutate(
    prev_arr_delay = lag(arr_delay),
    prev_arr_ts    = lag(arr_ts),
    time_since_prev = as.numeric(difftime(dep_ts, prev_arr_ts, units = "hours")),
    prev_flight_late = prev_arr_delay > 15 & time_since_prev <= 8
  ) %>%
  ungroup() %>%
  mutate(
    is_dep_delay = factor(dep_delay > 15, levels = c(FALSE, TRUE)),
    is_arr_delay = factor(arr_delay > 15, levels = c(FALSE, TRUE)),
    hour_of_day = dep_time %/% 100,
    # Fallback to lubridate if day_of_week column is absent
    is_weekend = day_of_week %in% c("6", "7"),
    ) %>%
  group_by(hour_of_day) %>%
  mutate(congestion = n()) %>%
  ungroup %>%
  filter(cancelled == '0')




# Define the 5-fold Cross-Validation strategy
set.seed(42)
train_control <- trainControl(method = "cv", number = 5)
```

## Models
### Departure Delay

```{r}
#| eval: false
# Limit data to relevant variables
dep_flight_filtered <- flights_filtered %>%
  filter(cancelled == '0') %>%
  filter(origin == "BWI", dest == "JFK") %>%
  select(c("dep_delay","is_dep_delay","prev_flight_late","dep_tmpf","dep_relh","dep_vsby","reporting_airline","dep_time_blk","is_weekend")) %>%
  # Scale weather vars
  mutate(across(all_of(c("dep_tmpf","dep_relh","dep_vsby")), ~ as.numeric(scale(.))))


dep_flight <- dep_flight_filtered %>%
  filter(
    !is.na(dep_delay),
    !is.na(prev_flight_late),    
    dep_delay >= -60,
    dep_delay <= 1440

  )

```

#### Regression
##### Linear

```{r, warning = FALSE}
#| eval: false
model_lm <- train(
  dep_delay ~ prev_flight_late + dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "lm", trControl = train_control
)

interpret_lm <- function(model) {
  tidy(model$finalModel) %>%
    mutate(
      abs_estimate = abs(estimate)
    ) %>%
    arrange(desc(abs_estimate))
}

interpret_lm(model_lm)
```

###### Model Ridge

```{r}
#| eval: false
model_ridge <- train(
  dep_delay ~ prev_flight_late + dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)

interpret_glmnet <- function(model) {
  lambda <- model$bestTune$lambda
  coefs <- coef(model$finalModel, s = lambda)
  
  tibble(
    variable = rownames(coefs),
    coefficient = as.numeric(coefs)
  ) %>%
    filter(variable != "(Intercept)") %>%
    mutate(abs_coef = abs(coefficient)) %>%
    arrange(desc(abs_coef))
}

interpret_glmnet(model_ridge)
```

###### Lasso

```{r}
#| eval: false
model_lasso <- train(
  dep_delay ~ prev_flight_late + dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)
interpret_glmnet(model_lasso)
```

##### Elastic Net

```{r}
#| eval: false
model_enet <- train(
  dep_delay ~ prev_flight_late + dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glmnet",
  trControl = train_control
)
interpret_glmnet(model_enet)
```

#### Classification
##### Logistic Regression

```{r}
#| eval: false
model_logit <- train(
  is_dep_delay ~ prev_flight_late + dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)
summary(model_logit)
```

##### PCA

```{r}
#| eval: false
# Select numeric columns; drop zero-variance and NA rows
pca_input <- dep_flight %>%
  select(dep_delay, dep_tmpf, dep_relh, dep_vsby) %>%
  select(where(~ n_distinct(.) > 1)) %>%
  drop_na()

pca_results <- prcomp(pca_input, scale. = TRUE)
summary(pca_results)
plot(pca_results, type = "l", main = "PCA Variance Explained")
biplot(pca_results, cex = 0.7, main = "PCA: Flight Feature Relationships")
```

### Model Comparison

```{r}
#| eval: false
results <- resamples(list(
  Linear     = model_lm,
  Ridge      = model_ridge,
  Lasso      = model_lasso,
  ElasticNet = model_enet
))

print(dotplot(results, metric = "Rsquared", main = "Model Comparison: R-squared"))
```

```{r}
#| eval: false
# Predicted vs Actual — Linear Regression
dep_flight$predictions <- predict(model_lm, newdata = dep_flight)

ggplot(dep_flight, aes(x = dep_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Departure Delay",
    x = "Actual Departure Delay (min)",
    y = "Predicted Departure Delay (min)"
  ) +
  theme_minimal()
```


```{r}
#| eval: false
# Confusion Matrix — Logistic Regression
logit_preds <- predict(model_logit, dep_flight)
confusionMatrix(logit_preds, dep_flight$is_dep_delay)
```

### Departure Delay where the Incoming Flight is not Delayed

```{r}
#| eval: false
# Filter to where incoming flight is not late
dep_flight <- dep_flight %>%
  filter(prev_flight_late == FALSE)
```

#### Linear Regression

```{r}
#| eval: false
model_lm <- train(
  dep_delay ~ dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "lm", trControl = train_control
)

interpret_lm(model_lm)

```

#### Ridge

```{r}
#| eval: false
model_ridge <- train(
  dep_delay ~ dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)

interpret_glmnet(model_ridge)
```

#### Lasso

```{r}
#| eval: false
model_lasso <- train(
  dep_delay ~ dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)
interpret_glmnet(model_lasso)

```

#### Elastic Net

```{r}
#| eval: false
model_enet <- train(
  dep_delay ~ dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glmnet",
  trControl = train_control
)
interpret_glmnet(model_enet)

```

#### Logistic Regression

```{r}
#| eval: false
model_logit <- train(
  is_dep_delay ~ dep_tmpf + dep_relh + dep_vsby + reporting_airline + dep_time_blk + is_weekend,
  data = dep_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)
summary(model_logit)
```

#### PCA

```{r}
#| eval: false
# Select numeric columns; drop zero-variance and NA rows
pca_input <- dep_flight %>%
  select(dep_delay, dep_tmpf, dep_relh, dep_vsby) %>%
  select(where(~ n_distinct(.) > 1)) %>%
  drop_na()

pca_results <- prcomp(pca_input, scale. = TRUE)
summary(pca_results)
plot(pca_results, type = "l", main = "PCA Variance Explained")
biplot(pca_results, cex = 0.7, main = "PCA: Flight Feature Relationships")
```

#### Model Result Comparison

```{r}
#| eval: false
results <- resamples(list(
  Linear     = model_lm,
  Ridge      = model_ridge,
  Lasso      = model_lasso,
  ElasticNet = model_enet
))

print(dotplot(results, metric = "Rsquared", main = "Model Comparison: R-squared"))
```

#### Plot Predicted vs Actual

```{r}
# Predicted vs Actual — Linear Regression
dep_flight$predictions <- predict(model_lm, newdata = dep_flight)

ggplot(dep_flight, aes(x = dep_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Departure Delay - Where Previous Flight was not Delayed",
    x = "Actual Departure Delay (min)",
    y = "Predicted Departure Delay (min)"
  ) +
  theme_minimal()
```

#### Confusion Matrix

```{r}
# Confusion Matrix — Logistic Regression
logit_preds <- predict(model_logit, dep_flight)
confusionMatrix(logit_preds, dep_flight$is_dep_delay)
```


### Arrival Delay

```{r}
# Limit data to relevant variables
arr_flight_filtered <- flights_filtered %>%
  filter(cancelled == '0') %>%
  filter(origin == "JFK", dest == "BWI") %>%
  select(c("arr_delay","is_arr_delay","is_dep_delay","arr_tmpf","arr_relh","arr_vsby","reporting_airline","arr_time_blk","is_weekend")) %>%
  # Scale weather vars
  mutate(across(all_of(c("arr_tmpf","arr_relh","arr_vsby")), ~ as.numeric(scale(.))))


arr_flight <- arr_flight_filtered %>%
  filter(
    !is.na(arr_delay),
    !is.na(is_dep_delay),    
    arr_delay >= -60,
    arr_delay <= 1440

  )

```

#### Regression
##### Linear

```{r}
model_lm <- train(
  arr_delay ~ is_dep_delay + arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "lm", trControl = train_control
)
interpret_lm(model_lm)
```


###### Model Ridge

```{r}
model_ridge <- train(
  arr_delay ~ is_dep_delay + arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)
interpret_glmnet(model_ridge)
```


###### Lasso

```{r}
model_lasso <- train(
  arr_delay ~ is_dep_delay + arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)
interpret_glmnet(model_lasso)
```


##### Elastic Net

```{r}
model_enet <- train(
  arr_delay ~ is_dep_delay + arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glmnet",
  trControl = train_control
)
interpret_glmnet(model_enet)
```


#### Classification
##### Logistic Regression

```{r}
model_logit <- train(
  is_arr_delay ~ is_dep_delay + arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)
summary(model_logit)
```


##### PCA

```{r}
# Select numeric columns; drop zero-variance and NA rows
pca_input <- arr_flight %>%
  select(arr_delay, arr_tmpf, arr_relh, arr_vsby) %>%
  select(where(~ n_distinct(.) > 1)) %>%
  drop_na()

pca_results <- prcomp(pca_input, scale. = TRUE)
summary(pca_results)
plot(pca_results, type = "l", main = "PCA Variance Explained")
biplot(pca_results, cex = 0.7, main = "PCA: Flight Feature Relationships")
```


### Model Comparison

```{r}
results <- resamples(list(
  Linear     = model_lm,
  Ridge      = model_ridge,
  Lasso      = model_lasso,
  ElasticNet = model_enet
))

print(dotplot(results, metric = "Rsquared", main = "Model Comparison: R-squared"))
```


```{r}
# Predicted vs Actual — Linear Regression
arr_flight$predictions <- predict(model_lm, newdata = arr_flight)

ggplot(arr_flight, aes(x = arr_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Arrival Delay",
    x = "Actual Arrival Delay (min)",
    y = "Predicted Arrival Delay (min)"
  ) +
  theme_minimal()
```



```{r}
# Confusion Matrix — Logistic Regression
logit_preds <- predict(model_logit, arr_flight)
confusionMatrix(logit_preds, arr_flight$is_arr_delay)
```

### Arrival Delay where the Flight Departure is not Delayed

```{r}
# Filter to where flight is not late upon departure
arr_flight <- arr_flight %>%
  filter(is_dep_delay == FALSE)
```

#### Linear Regression

```{r}
model_lm <- train(
  arr_delay ~ arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "lm", trControl = train_control
)
interpret_lm(model_lm)
```

#### Ridge

```{r}
model_ridge <- train(
  arr_delay ~ arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)
interpret_glmnet(model_ridge)
```

#### Lasso

```{r}
model_lasso <- train(
  arr_delay ~ arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glmnet",
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, by = 0.1)),
  trControl = train_control
)
interpret_glmnet(model_lasso)
```

#### Elastic Net

```{r}
model_enet <- train(
  arr_delay ~ arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glmnet",
  trControl = train_control
)
interpret_glmnet(model_enet)
```

#### Logistic Regression

```{r}
model_logit <- train(
  is_arr_delay ~ arr_tmpf + arr_relh + arr_vsby + reporting_airline + arr_time_blk + is_weekend,
  data = arr_flight, method = "glm",
  family = "binomial",
  trControl = train_control
)
summary(model_logit)
```

#### PCA

```{r}
# Select numeric columns; drop zero-variance and NA rows
pca_input <- arr_flight %>%
  select(arr_delay, arr_tmpf, arr_relh, arr_vsby) %>%
  select(where(~ n_distinct(.) > 1)) %>%
  drop_na()

pca_results <- prcomp(pca_input, scale. = TRUE)
summary(pca_results)
plot(pca_results, type = "l", main = "PCA Variance Explained")
biplot(pca_results, cex = 0.7, main = "PCA: Flight Feature Relationships")
```

#### Model Performance Comparison

```{r}
results <- resamples(list(
  Linear     = model_lm,
  Ridge      = model_ridge,
  Lasso      = model_lasso,
  ElasticNet = model_enet
))

print(dotplot(results, metric = "Rsquared", main = "Model Comparison: R-squared"))
```

#### Predicted vs Actual

```{r}
# Predicted vs Actual — Linear Regression
arr_flight$predictions <- predict(model_lm, newdata = arr_flight)

ggplot(arr_flight, aes(x = arr_delay, y = predictions)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Linear Regression: Actual vs. Predicted Arrival Delay - Where Flight was not Late upon Departure",
    x = "Actual Arrival Delay (min)",
    y = "Predicted Arrival Delay (min)"
  ) +
  theme_minimal()
```

#### Confusion Matrix

```{r}
# Confusion Matrix — Logistic Regression
logit_preds <- predict(model_logit, arr_flight)
confusionMatrix(logit_preds, arr_flight$is_arr_delay)
```